### 2019-05-20 Writing/response_to_reviewers.tex

\item{\textcolor{blue}{I have some concerns about the approach, however. I was somewhat confused by the CD metric. When quantifying n, how is it decided how many taxa have "comparable" characters?
If you simulated the character for all taxa, do they not all have this character (i.e. n always equals the dataset size)?
But more importantly, will two characters not be more similar if they share the same number of character states, independent of if they are telling you the same evolutionary history?
It seems to me that whether or not two characters share a state space is not the relevant axis along which to measure correlation.}}
\label{gower}

Using a Gower based distance, we only compared the comparable taxa between characters.
For example, for two characters \texttt{x = {0, 1, 2, 3}} and \texttt{y = {0, ?, ?, ?}}, the Character Difference metric will be effectively calculated using only the first taxon (\texttt{x = {0}} and \texttt{y = {0}}) and would result in a Character Difference of 0.
As the reviewer points out, characters with less taxa in common will thus be less likely to be different.
This is consistent with what the CD metric actually captures in terms of phylogenetic signal (the number of share splits).
In the example above, the both characters imply at least one split (between the first taxon and the others).
In other words, in a phylogenetic software, both characters will support the same split.

As the reviewer also correctly points out, we simulated our data with no missing data so this does not affect our results.
The precision in the manuscript was just describing the properties of the new CD metric.

We've clarified how missing data can be potentially treated in the methods section:
\textit{For example, for two characters X and Y for four taxa, if character \texttt{X = \{0,0,?,1\}} for the first, second, third and fourth taxa respectively and character \texttt{Y = \{0,0,1,1\}}, only the first, second and fourth taxa are compared for this character.
Note that in these simulations, no missing data was included so the characters are compared for equal number of taxa (see below).} lines 213-218.

@@

We've added the proportion of replaced (i.e. duplicated) characters in each matrices for each ``minimised'', ``maximised'' and ``randomised'' scenarios along with the changes in overall Character Differences in the supplementary material 3, figures 4, 5 and 6.

The maximising/minimising character differences was done \textit{a priori} without any targeted number of characters to replace to make the analysis less idiosynchratic.
In other words, we decided to modify the randomly simulated matrices to globally increase/decrease character differences which is expected to decrease anyway in bigger matrices.
When using a small number of character states (i.e. two or three), the matrices with few taxa and many characters are expected to have an overall low character difference because of the smaller combinatoric possibilities for each characters.

We've added mentions to the proportion of replaced characters analysis in the supplementary and added a paragraph explaining this caveat in the discussion:

\textit{Furthermore, it is important to note that on average, less characters where duplicated (replace) in the ``minimised'' matrices compared to both the ``maximised'' and ``randomised'' matrices (supplementary material 3, Figs 4, 5 and 6).
However, this change in the amount of replaced characters did not changed the way average character differences changed across the three transformed matrices: the character differences were always consistently increased (``maximised''), decreased (``minimised'') and maintained at a normal level (``randomised'') for each replicate (supplementary material 3, Figs 1, 2 and 3).
This suggest that character difference can be decreased with a smaller amount of characters replacements than increasing it.} lines 453-462.

@@

\textit{For example, two characters ``tail colour'' and ``tail length'' could be coded two times as an absence for a taxon with no tail.} lines 95-96 and \textit{In the example above, attributes of the tail (colour or length) are hierarchically dependent on the presence of the tail in the first place.} lines 102-104.

\item{\textcolor{blue}{"Correlation between characters detected by the software:"\\
I really struggled with this section. It seems like the problem here is still a coding problem - that the authors of these four characters have chosen four highly dependent characters to build the matrix. Even scoring by hand under parsimony would have problems here. I think what you need to decide here is what you want to communicate in this paragraph. Is it that software is not currently equipped to handle this issue?}}

We've added a paragraph highlighting how the problem indeed comes from the fact that current software are not equipped to handle this issue:

\textit{In fact, for practicality reasons, phylogenetic software do not (or rarely) distinguish between the different sources of correlation.
In this specific example, we would intuitively want to favour character \textbf{C1} and \textbf{C2} that seem to be more likely to reflect the split by separating artiodactyles from the cetaceans.
In other words, this is the correlation between characters that depicts the evolutionary history of this group.
However, the other characters depict the same history (i.e. the same split) but for the wrong reasons (note that even if \textbf{C2} was coded presence (1) or absence (0) of baleen, the same split would still be estimated by the software).
One solution would be to a) weight some characters (e.g. giving \textbf{C3} and \textbf{C4} a lower weight) and b) make some characters dependent on each other (e.g. making \textbf{C3} and \textbf{C4} dependent on \textbf{C1}).
However, this can also be a bad practice since it can easily lead to the resulting tree reflecting the researcher's prior expectations rather than the true evolutionary history.} lines 145-157.

@@

We've clarified this sentence (see Reviewer 1, Major suggestion \ref{gower}).

\item{\textcolor{blue}{"This procedure was used to treat all the characters are unordered with no assumption on the meaning of the character state"\\
Why is this important? The text doesn't mention testing anything related to orderedness, or specifying anything related to orderedness in MrBayes.}}

We removed the mention of orderdness of the characters since, indeed, it might be more confusing than helpful in this context (we didn't not simulate any ordered characters and the CD metric does not consider characters as ordered - see supplementary material 1). We've changed the sentence to the following:

\textit{This procedure allows to compare characters regardless of the significance given to their tokens (following the \textit{xyz} notation in Felsenstein, 2004; as used in D\'{a}valos et al., 2014).} lines 221-223.

@@

\item{\textcolor{blue}{"The morphological HKY model"\\
Is not real. This is just the HKY model. Goloboff (2017) would dispute that this model does not favor a Bayesian method, since it is still a Markovian model. That O'Reilly et al. used this model to generate binary data does not mean one has to - you could translate each of the four nucleotides to its own discrete character. This would probably get even further from the Mk model assumptions than collapsing the multistate character produced via the HKY model to a binary character.}}

We've enquoted the ``morphological HKY model'' and specified that it is still a Markovian model:
\textit{Note that both models (``morphological HKY'' and Mk) are both based on a Markovian model and only differ in the number of states generated and transformed.
However, the ``morphological HKY model'' is not the one} directly \textit{estimated in the phylogenetic software used in this protocol (MrBayes, see below).} lines 271-275.

@@

We meant that modifying phylogenetic matrices after the phylogenetic inference is a way to test the effect of the sources of correlation but is probably not a good practice in systematics.
Although re-weighting characters based on consistency or retention indices can be used to objectively improve the topology, changing the matrix's characters \textit{a posteriori}, it can be seen as dishonest to remove characters because they ``don't give the right signal''). We've removed the end of this sentence however to avoid any further confusion.

\item{\textcolor{blue}{"However, we do not compare the ``maximised'', ``minimised'' and ``randomised'' to the ``true'' tree but rather to the ``normal'' tree. "\\
I understand why you did this, but I think you should also compare normal to true. That will give us some idea of the scope of how close we can get to true, under these conditions, and will help contextualize the results.}}

We've added the comparisons between the ``normal'' and ``randomised'' trees to the ``true'' tree in the supplementary material 3, figures 9 to 11.
We've added a reference to these figures in the main text:

\textit{However, we did report the comparison between the ``true'' trees and both the ``normal'' and ``randomised'' in the supplementary materials 3, Figs 12, 13 and 14.} lines 346-347

@@

We've correct the supplementary materials 1 and added more explanations on (1) the character translation, (2) the four-point metric proof and (3) on how to interpret the $CD$ metric.
Additionally, we added more precisions in the text:

\textit{(i.e. entirely correlated characters would define identical splits between a set of taxa)}. lines 202-203.

\item{\textcolor{blue}{While some details of the simulation study are included, some important ones are missing. In particular, in the design of the study, how did you avoid or mitigate errors introduced by the programs simulating data? All of the programs use random-number generators and have their own biases. How have you accounted for this in the study of design to conclude that your results are due to signal, and not bias from the programs used?}}

We do agree that generating true randomness in current software is tricky but we are not sure how to answer this comment.
The three first steps (generating the tree, the matrices and modifying the matrices) were using the default random number generator implemented in R, the Mersenne Twister algorithm that has a cycle of $2^{32}-1 = 4.2e^{9}$ unsigned random integers (see random numbers documentation in R: \texttt{?.Random.seed} - we used the implementation from \texttt{R} v3.5 that differs in the starting seed from the new \texttt{R} v3.6).
The last step (generating the trees) was using both the random number generated by default my MrBayes and PAUP* based on the seeds from the different HPC we used (the Imperial College London Cluster Services and the Australian QRIScloud Awoonga cluster).

@@

Therefore, if cumulating all the seeds used in the generation of the matrices (trees + characters) and the modifications of the matrices is likely to be no more than $6.4e^{7}$.
Therefore we do not believe that the random number generation was a proper since we sampled two orders of magnitude less random numbers than the total ones available.

\item{\textcolor{blue}{How did you ``randomly'' modify your matrices? Did you sample under some distribution? If so, which one?}}

Every replaced characters were randomly selected randomly from the total sample of characters (i.e. randomly sampling from the uniform integers distribution (1, $c$); where $c$ is the number of remaining characters). We have now specified this in the text:

\textit{Each replaced characters (i.e. the characters with a weight of +$1$) were randomly selected by sampling them randomly from the list of the remaining characters.} lines 297-299.

@@

There is actually a vigorous debate surrounding how to generate biologically meaningful morphological characters \citep[][]{OReilly20160081,goloboff2017weighted,puttick2017uncertain,GoloboffEmpirical,OReillyEmpirical}.
This ongoing debate, however, focuses mainly on inference methods where data simulation is a tangential aspect of it.
Therefore, we decided to base our simulation protocol on the empirical observations showing that half of 206 analysed datasets supported a M$k$-like model \citep[][Fig. 3]{Wright01072016} and the other half being a combination of models not explicitly described here but that are believed to be covered with the HKY-binarised model \citep{puttick2017uncertain,OReillyEmpirical}.
In our simulations, the idea was to cover a broad range of morphological characters (with variable state numbers, evolution and transition rates) in order to capture a realistic image of the ``discrete morphological character space''.

\item{\textcolor{blue}{Similarly, why did you choose 0.85 for the Mk model? Does it represent some important set of character sequences? How does it represent the space in general?}}

This 0.85 proportion of binary characters and 0.15 of three state ones was extracted from the 100 empirical matrices analysed in \cite{Guillerme2016146} (Supplementary appendix 1. Appendix A: Tree inference) and \cite{ZouConvergence}.
We've specified this in the text:

\textit{We draw the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters (from empirical observations in: Guillerme and Cooper, 2016b; Zou and Zhang, 2016).} lines 256-258.

@@

\subsection{Major suggestions}

\begin{enumerate}

\item{\textcolor{blue}{I feel that the manuscript is certainly worth publishing in Systematic Biology, although I worry that the simulation procedure used by the authors limits their ability to thoroughly test of the effect of between character correlation on the accuracy of estimating phylogenetic trees. The authors note this themselves in the discussion, but I feel more should be made of this point. If the simulations utilised an explicit distribution of character covariance then I think that interpretation of the results would have been more straightforward.}}
\label{stramlined}

As noted in the introduction, the aim of this manuscript was mainly to see does different levels of correlation between discrete morphological characters affect tree topology.
We also review different sources of correlation for characters and discuss how the multidimensional complexity of a phenotype (and the correlations therein) is not reflected in two dimensional character matrices.
We do agree with the reviewer that a thorough analysis of the variance/covariance structure of simulated ``discretised'' characters (see reviewer 3, major suggestion \ref{hiden}) would have broad more insight into the nature of character correlation on phylogenetics inference in general (regardless of the type of character and correlation) but due to the length required to run the current ``simplistic'' simulation, we feel that this would be out of the scope of this manuscript.

We have now clarified in the manuscript what our objectives were by streamlining the last paragraph of the introduction (lines 160-168).
We have also explicitly added in the supplementary material 3, figs 1, 2 and 3 how the character difference changes between all matrices compared to the ``normal'' matrix.

@@

\item{\textcolor{blue}{Another worry I have is that the implied covariance at the tips does not necessarily reflect covariation over the full phylogeny. It is conceivable that two characters can appear to have near perfect correlation at the tips, but arrived at these states through completely independent pathways, particularly when simulating with a continuous time Markov chain where unobserved transitions to a new state and back to the original state are possible.}}
\label{hiden}

We agree with this reviewer's comment which can be especially thought to be the case for discrete characters with a fast rate of evolution.
Such characters will likely swap between states along the branches following a hidden Markov chain.

However, we feel that modelling this process might not be useful in this specific study:
First, in empirical data two characters can be strongly correlated for biological and evolutionary reasons (e.g. the absence of an astragalus and the presence of baleen in table 1) leading to a strong correlation of both characters at the tips (or even nodes) but does not mean that both character's hidden Markov chain are correlated.
Second, by comparing characters at any time point rather than simply at the tips (which can be problematic since our trees were not all ultrametric) we would probably not be able to compare the effect of the ``true'' characters' correlation since this information will not be used by the phylogenetic software (MrBayes and PAUP*) that will only use tip values.

We've added this remark in the caveats section:
\textit{It is also worth noting that character similarity at the tips does not necessarily reflects that the characters underwent the same evolutionary history.
This may be specifically true when simulating characters based on the M$k$ model where the resulting tip states does not necessarily reflects variation of the state along the branches (e.g. Revell, 2014). However, it is worth noting that in empirical studies, such state changes along the branches are nearly always empirically unknown (even though they can be estimated by Maximum Likelihood).} lines 554-560.

@@

\item{\textcolor{blue}{It seems that the results reported by the authors are inherently driven by the phylogenetic signal present in their ``bootstrapped'' datasets. This effect would be present in any approach to the simulation of covarying characters, but as the authors do not explicitly simulate covariance but instead rely on a proxy they cannot be certain that their results are driven by true character covariation or purely a restructuring of phylogenetic signal by resampling. In a simulation framework in which covariance is actively enforced and controlled it would be possible to assess other aspects of character covariance that seem important for empirical datasets, such as correlation within morphological compartments. I think without tests of these sort we cannot be completely sure of the effect of character covariance on topology estimates. Despite this, the authors present an overview of the effect of character covariance at large, and I certainly think that this is a worthwhile contribution.}}

We do agree with this reviewer's comment and believe that a study looking at difference in covariation levels between character in different morphological modules would be really useful (e.g. are cranial character more correlated than postcranial)?
In our simulation protocol we effectively only simulated characters from a single ``module'' since they were generated based on the same topology (and branch length) and the same distribution (rates where sampled from a single distribution).
Therefore, we believe that the correlation approximated by the $CD$ metric was actually explicitly simulated through our protocol.
This is also reflected by the fact that, in practice, when increasing the correlation (reducing character difference), rather few characters needed to be duplicated (see supplementary material 3, figures 1 to 6).

We hope to expand this study in the future by create different groups of characters sharing the same overall correlation by generating ``bunches'' of characters based on different rates (to simulate rate homogeneity) and different trees (branch length and topology - to simulate variation in ``gene'' trees).

@@

\item{\textcolor{blue}{A discussion of the ``effective'' number of characters might also be warranted. When maximising character covariance this likely requires the resampling of the same handful of characters multiple times, each of which provides the exact same phylogenetic signal. The redundancy of these characters seems likely to influence the accuracy of estimates. The choice of characters that are sampled by the bootstrap procedure will therefore constrain the accuracy of topology estimates. If the resampled matrix consists of a handful of characters with high consistency with the true tree, sampled many times over, then the tree estimate will be good (for at least part of the tree) and covariance will be high. Similarly, if the handful of regularly resampled characters are mainly inconsistent (with the true tree) then tree estimates will be poor but the implied covariance will also be high. I think this is evident from the reduced performance in larger matrices, where there will be a significant reduction in phylogenetic signal when attempting to maximise covariance.}}

We've added the following part to the discussion:

\textit{Overall, it is expected that minimising character difference (increasing character correlation) is likely to have a bigger effect on recovering the ``correct'' topology since most characters will provide the same phylogenetic signal.
If the ``minimised'' matrix consists of a handful of characters with high consistency with the ``best'' tree, then the tree estimate will be good (for at least part of the tree).
Similarly, the characters in the ``minimised'' matrix are mainly inconsistent with the ``best'' tree then the estimates will be poor.} lines 468-474

### 2019-05-20 Writing/draft_highlighted.tex
finished updated response

\subsection{Simulating discrete morphological matrices}
To simulate the matrices we applied a protocol very similar to \cite{Guillerme2016146}.
First, we generate random birth-death trees with the birth ($\lambda$) and death ($\mu$) parameters sampled from a uniform $(0,1)$ distribution maintaining $\lambda$ $>$ $\mu$ using the \texttt{diversitree} \texttt{R} package \citep[v0.9-8;][]{fitzjohndiversitree2012} and saving the tree after reaching either 25, 75 or 150 taxa.
For each tree, we arbitrarily set the outgroup to be the first taxon (alphabetically) thus effectively rooting the trees on this taxon.
These trees are hereafter called the ``true'' trees (see distinction below).
We then simulated discrete morphological characters on the topology of these trees using the either of the two following models:
\begin{itemize}
    \item The \hl{``morphological HKY-binary'' model} \citep{OReilly20160081} which is an HKY model \citep{HKY85} with a random states frequency (sampled from a Dirichlet distribution $Dir(1,1,1,1)$) and using a transition/transvertion rate of $2$ \citep{douadycomparison2003} but where the purines (A,G) were changed into state $0$ and the pyrimidines (C,T) in state $1$.
    This model has the advantage of not favouring Bayesian inference \citep[since it doesn't use an M$k$ model;][; see discussion]{OReilly20160081} but the downside of it is it can only generate binary state characters \citep[or 4 states;][]{puttick2017uncertain}.
    \item To generate more than binary states characters, we used the M$k$ model \citep{lewisa2001}.
    We draw the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters \citep[\hl{from empirical observations in: }][]{Guillerme2016146,ZouConvergence}.
    This model assumes a equal transition rate between character states which might seem overly simplistic, excluding other observed transition patterns \citep[e.g. Dollo characters;][]{Dollo,wright2015came}.
    However, \cite{Wright01072016} \hl{have shown that equal transition rate is a reasonable assumption for many empirical datasets.}
\end{itemize}

@@

\subsection{Limitations}
First, simulating evolutionary history is complex.
Not only because the models we're using to infer phylogenies are ever improving \citep[e.g.][]{heath2014fossilized,Wright01072016} but also because generalising morphological evolution across vastly different organisms is probably impossible \citep[see constrasted discussions from][]{GoloboffEmpirical,OReillyEmpirical}.
However, we do not compare the ``maximised'', ``minimised'' and ``randomised'' to the ``true'' tree but rather to the ``normal'' tree.
This allows us to reduce the caveats from our simulations on the effect of character correlation since we only compare the simulation end products to each other (the outputs) rather than to the simulation inputs.
\hl{It is also worth noting that character similarity at the tips does not necessarily reflects that the characters underwent the same evolutionary history.
This can specifically true when simulating characters based on the M$k$ model where the resulting tip states does not necessarily reflects variation of the state along the branches}
 \citep[e.g.][]{revell2014ancestral}.
\hl{However, it is worth noting that in empirical studies, such state changes along the branches are nearly always unknown.}

>>>>

\subsection{Simulating discrete morphological matrices}
To simulate the matrices we applied a protocol very similar to \cite{Guillerme2016146}.
First, we generate random birth-death trees with the birth ($\lambda$) and death ($\mu$) parameters sampled from a uniform $(0,1)$ distribution maintaining $\lambda$ $>$ $\mu$ using the \texttt{diversitree} \texttt{R} package \citep[v0.9-8;][]{fitzjohndiversitree2012} and saving the tree after reaching either 25, 75 or 150 taxa.
For each tree, we arbitrarily set the outgroup to be the first taxon (alphabetically) thus effectively rooting the trees on this taxon.
These trees are hereafter called the ``true'' trees (see distinction below).
We then simulated discrete morphological characters on the topology of these trees using the either of the two following models:
\begin{itemize}
    \item The \hl{``morphological HKY-binary'' model} \citep{OReilly20160081} which is an HKY model \citep{HKY85} with a random states frequency (sampled from a Dirichlet distribution $Dir(1,1,1,1)$) and using a transition/transvertion rate of $2$ \citep{douadycomparison2003} but where the purines (A,G) were changed into state $0$ and the pyrimidines (C,T) in state $1$.
    This model has the advantage of not favouring Bayesian inference \citep[since it doesn't use an M$k$ model;][; see discussion]{OReilly20160081} but the downside of it is it can only generate binary state characters \citep[or 4 states;][]{puttick2017uncertain}.
    \item To generate more than binary states characters, we used the M$k$ model \citep{lewisa2001}.
    \hl{We draw the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters} \citep[\hl{from empirical observations in: }][]{Guillerme2016146,ZouConvergence}.
    This model assumes a equal transition rate between character states which might seem overly simplistic, excluding other observed transition patterns \citep[e.g. Dollo characters;][]{Dollo,wright2015came}.
    However, \cite{Wright01072016} \hl{have shown that equal transition rate is a reasonable assumption for many empirical datasets.}
\end{itemize}

@@

\subsection{Limitations}
First, simulating evolutionary history is complex.
Not only because the models we're using to infer phylogenies are ever improving \citep[e.g.][]{heath2014fossilized,Wright01072016} but also because generalising morphological evolution across vastly different organisms is probably impossible \citep[see constrasted discussions from][]{GoloboffEmpirical,OReillyEmpirical}.
However, we do not compare the ``maximised'', ``minimised'' and ``randomised'' to the ``true'' tree but rather to the ``normal'' tree.
This allows us to reduce the caveats from our simulations on the effect of character correlation since we only compare the simulation end products to each other (the outputs) rather than to the simulation inputs.
\hl{It is also worth noting that character similarity at the tips does not necessarily reflects that the characters underwent the same evolutionary history.
This may be specifically true when simulating characters based on the M$k$ model where the resulting tip states does not necessarily reflects variation of the state along the branches}
\citep[e.g.][]{revell2014ancestral}.
\hl{However, it is worth noting that in empirical studies, such state changes along the branches are nearly always empirically unknown (even though they can be estimated by Maximum Likelihood).}
